---
title: "Analyzing the Presidential Elections"
output: html_notebook
---

Adapted from _"Programming Skills for Data Science: Start Writing Code to Wrangle, Analyze, and Visualize Data with R, First Edition."_ by Freeman and Ross.

Load the `tidyverse`

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

Loading the presidential dataset into `presidentialElections` variable:

```{r}
# read data
presidentialElections <- read_csv("https://raw.githubusercontent.com/reisanar/datasets/master/presidentialElections.csv")
```


Select `year` and `demVotes` (percentage of vote won by the Democrat) from the `presidentialElections` data frame.

```{r}
votes <- presidentialElections %>%
  select(year, demVote)
votes
```

Select columns `state` through `year` (i.e., `state`, `demVote`, and `year`)

```{r}
presidentialElections %>%
  select(state:year)
```

Select all columns except for `south`

```{r}
presidentialElections %>%
  select(-south)
```


Select all rows from the 2008 election

```{r}
votes_2008 <- presidentialElections %>%
  filter(year == 2008)
votes_2008
```


Extract the row(s) for the state of Colorado in 2008

```{r}
votes_colorado_2008 <- presidentialElections %>%
  filter(year == 2008, state == "Colorado")
votes_colorado_2008
```

Add an `other_parties_vote` column that is the percentage of votes for other parties. Also add an `abs_vote_difference` column of the absolute difference between percentages. Note you can use columns as you create them!

```{r}
presidentialElections <- presidentialElections %>%
  mutate(
    other_parties_vote = 100 - demVote,
    abs_vote_difference = abs(demVote - other_parties_vote)
  )
presidentialElections
  
```

Arrange rows in decreasing order by `year`, then by `demVote` within each `year`

```{r}
presidentialElections <- presidentialElections %>%
                          arrange(desc(year), demVote)
presidentialElections
```

Compute summary statistics for the `presidentialElections` data frame

```{r}
average_votes <- presidentialElections %>%
  summarize(
    mean_dem_vote = mean(demVote),
    mean_other_parties = mean(other_parties_vote)
  )
average_votes
  
```


## Additional Analysis

_Which state had the highest percentage of votes for the Democratic Party candidate (Barack Obama) in 2008?_

Answering this seemingly simple question requires a few steps:

1. Filter down the data set to only observations from 2008.

2. Of the percentages in 2008, filter down to the one with the highest percentage of votes for a Democrat.

3. Select the name of the state that meets the above criteria.



Use the pipe operator to connect multiple instructions:


```{r}
presidentialElections %>%
  filter(year == 2008) %>%
  filter(demVote == max(demVote)) %>%
  select(state)


```


Compute summary statistics by state: average percentages across the years


