---
title: "Exploring Hollywood Movies with `dplyr`"
output: html_notebook
---


## Load packages

We will be using the set of tools for data transformation from the `dplyr` packages, part of the `tidyverse`. Load the `tidyverse` by executing the command below

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```


## Example: Hollywood Movies

Read data on Hollywood Movies

```{r}
mov_url <- "https://raw.githubusercontent.com/reisanar/datasets/master/HollywoodMovies.csv"
Hollywood <- read_csv(mov_url)
```

Learn more about the dataset by using the function `glimpse()`

```{r}
glimpse(Hollywood)
```


Print the first rows of the dataset 

```{r}
head(Hollywood, 10)
```


## `filter()`


- Filter all Universal movies

```{r}
filter(Hollywood, LeadStudio == "Universal")
```


- Get only action movies by Universal 

```{r}
filter(Hollywood, LeadStudio == "Universal", Genre == "Action")
```

- Get movies by Disney and store them in an object called `disney_movies`

```{r}
disney_movies <- Hollywood %>%
  filter(LeadStudio == "Disney")
  
```

```{r}
disney_movies
```


- What Disney movies are from 2010?

```{r}
disney_2010 <- disney_movies %>%
  filter(Year == "2010")
disney_2010
```


- Find Disney movies that had `Profitability` larger than 1500

```{r}
disney_movies %>%
  filter(Profitability > 1500)
```


## `mutate()`

Create a new variable called `Score` that is the average score of `RottenTomatoes` and `AudienceScore`

```{r}
mutate(Hollywood, Score = (RottenTomatoes + AudienceScore) / 2)
```


## `select()`

Create a new data frame that only contains columns: `Movie`, `Studio`, `Genre`, `Budget`, `Profitability`, and `Year`

```{r}
my_movies <- select(Hollywood, Movie, LeadStudio, Genre, Budget, Profitability, Year)
# print the new data frame
my_movies
```


## `arrange()`

Sort your new data frame based on the `Budget` variable

```{r}
my_movies %>%
  arrange(-Budget)
```


## `summarise()`

Get a summary of the average Budget for the movies in the `Hollywood` dataset

```{r}
summarise(Hollywood,
          ave_budget = mean(Budget, na.rm = TRUE))
```

Find the movie with the maximum budget:

```{r}
filter(Hollywood, Budget == max(Budget, na.rm = TRUE))
```

Find the movie with the minimum budget:

```{r}
filter(Hollywood, Budget == min(Budget, na.rm = TRUE))
```

Notice that in this dataset the movie "Soul Men" has a reported value of 0 for `Budget`. This might be due an error when collecting the records, or a (wrong) decision of assigning the value of 0 for a movie for which there was no record of budget (`NA` should have been the right choice in such case)


## `group_by()`


Create a summary of the average budget for movies per studio. Notice that we should use the pipe operator `%>%` for this task

```{r}
Hollywood %>%
  group_by(LeadStudio) %>%
  summarise(AveBudget = mean(Budget, na.rm = TRUE))

```

Arrange the results from the previous step in descending order

```{r}
Hollywood %>%
  group_by(LeadStudio) %>%
  summarise(AveBudget = mean(Budget, na.rm = TRUE)) %>%
  arrange(-AveBudget)

```


## `rename()`

The function `rename()` works in the following way:

`rename(df, new_name = old_name)`

- Rename the variable `LeadStudio` for `Studio`

```{r}
rename(Hollywood, Studio = LeadStudio)
```



