---
title: "Blood Pressure and Steps - First Exploration"
output: html_notebook
---

## Load packages and read the data

Load the `tidyverse` package for data exploration

```{r}
library(tidyverse)
```


Read the data: 

```{r}
bp_steps <- read_csv("https://raw.githubusercontent.com/reisanar/datasets/master/bp_and_steps.csv")
```
Notice the _type_ of some of the columns in this dataframe. 


## Initial Exploration


Take a look at the dataset by printing the dataframe:

```{r}
bp_steps
```

- What initial questions can we formulate about this data? 

# Blood Pressure

1. Heart rate vs blood pressure? 

2. Relation of heart rate/bp vs time? 



Filter by vital categories

- How can we answer such questions?

1. ggplot via scatterplot "geom_point"

2. 

- Do all values make sense? Do you notice any _outliers_? 

```{r}
#heart rate vs blood pressure
bp_steps %>%
  filter(Vital == "Blood Pressure") %>%

ggplot() +
  geom_point(aes(x = HeartRate, y = BP_Sys), na.rm = TRUE)
```



## Data Dictionary


| Field Name  | Field Definition                                    |
| ------------|-----------------------------------------------------|
| `Vital`     | One of two: Blood Pressure, Steps                   |
| `Value`     | The collected value of the `Vital` column           |
| `HeartRate` | Average heart rate within a fifteen minute interval |
| `DateSplit` | The date the data is collected                      |                                             
| `TimeSplit` | The time the data is collected                      |                                             
| `BP_Sys`    | Systolic blood pressure reading                     |
| `BP_Dia`    | Diastolic blood pressure reading                    |



_Some notes:_

- Normal heart rate range for an adult is between 60-100 beats per minute. During strenuous exercise, it is normal for this to increase up to 170 beats per minute for individuals over 18. 

- Systolic blood pressure reading is an average over a fifteen minute time interval.  

- Diastolic blood pressure reading as an average over a fifteen minute time interval.    

