---
title: "Tidy up your data " 
author: "Data Science Student"
output: html_notebook
---

For this assignment you will be using the `gather()`, `spread()`, and `separate()` functions (along with function from `dplyr`) to _tidy up_ some data sets


## Problem 1


1. Load the `tidyverse` package

```{r}
library(tidyverse)
```

2. Consider the following data frame

```{r}
nba <- data.frame(
  player = c("James", "Durant", "Curry", "Harden", "Paul", "Irving"),
  team = c("CLEOH", "OAKCA", "OAKCA", "HOUTX", "HOUTX", "BOSMA"),
  day1points = c(25,23,30,41,26,20),
  day2points = c(24,25,33,45,26,23), row.names = NULL, stringsAsFactors = FALSE
)
nba
```


3. Use `gather()` to create a data frame with columns `day` and `points`

```{r}

```


4. Use `separate()` to create two columns: one for the "city" and the other for the "state" from the column "team". Set the `sep` option to 3. 

```{r}

```


5. Taking into account what you did in the previous exercises, use the right commands to _tidy up_ your dataset by creating 5 columns: "player", "team", "state", "day" and "points". 

```{r}

```


- Plot your dataset by creating a _scatterplot_ with `day` in the x-axis and `points` in the y-axis. Use `aes(color = city, shape = player)` as aesthetics.

```{r}
nba_tidy
```
```{r}

```




## Problem 2

1. Load the `tidyverse` package

```{r}
library(tidyverse)
```


2. Read the `grades204.csv` data set

```{r}
link <- "https://raw.githubusercontent.com/reisanar/datasets/master/grades204.csv"
mygrades <- read_csv(link)
```

3. Print the data set: it includes grades for 4 different quizzes, 4 exams, 3 HW assignments, and one final test

```{r}
mygrades
```

4. Separate the `name` column into `firstName` and `lastName`

```{r}

```

5. Create a column `username` separating the email handle from its domain

```{r}

```

6. The following code creates a column `ave.quiz` with the average quiz score for every entry:

```{r}

```

- Use the code above to create a column `ave.exam` in a similar way

```{r}

```

- Use the code above to create a column `ave.HW` in a similar way

```{r}

```


7. The following code shows an alternative procedure to compute the averages of quizzes, exams and assignments. Notice the use of the function `rowMeans()`

```{r}

```

- Using the example above, create a column called `total` that computes the total score based on the following point-breakdown:

20% Quizzes Average
15% HW Average
15% Final exam
50% Exams average

```{r}

```

Save this data frame into an object with the name `allgrades`

```{r}

```

8. Write a conditional statement that returns the following point breakdown letter grade:

Letter grade  | Total   
--------------|---------
A             | 90 - 100
B             | 80 - 89
C             | 67 - 79
D             | 60 - 66
F             | 0 - 59

```{r}

```

9. Use the conditional statement from the previous exercise to write a _function_ called `fgrade` that returns the letter grade (as defined before) based on the input score `total`

```{r}

```

Test your function with the scores 94, 85.3, and 57.3

```{r}

```

10. Use your function `fgrade` to add a final column to the data frame `allgrades`, with the final letter grade in each case. One way to do this is using the `map_chr()` function 

```{r}

```

