---
title: "Working with Vectors"
output: html_notebook
---

# Creating and Operating on Vectors

Create a vector `names` that contains your name and the names of 2 people 
next to you. Print the vector.

```{r}

names <- c("Gavin","Zachary","Patrick")
```


Use the colon operator `:` to create a vector `n` of numbers from 10 to 49

```{r}
n <- 10:49
```

Use the `length()` function to get the number of elements in `n`

```{r}
length(n)
```

Add 1 to each element in `n` and print the result

```{r}
print(n+1)
```

Create a vector `m` that contains the numbers 10 to 1 (in that order). Hint: use the `seq()` function

```{r}
m <- seq(10, 1)
```


Subtract `m` from `n`. Note the recycling!

```{r}
n_less_m <- n - m
print(n_less_m)
```


Use the `seq()` function to produce a range of numbers from -5 to 10 in `0.1` increments. Store it in a variable `x_range`

```{r}
x_range <- seq(-5,10, by = 0.1)
```


Create a vector `sin_wave` by calling the `sin()` function on each element in `x_range`.

```{r}
sin_wave <- sin(x_range)
```


Create a vector `cos_wave` by calling the `cos()` function on each element 
in `x_range`.

```{r}
cos_wave <- cos(x_range)
```


Create a vector `wave` by multiplying `sin_wave` and `cos_wave` together, then
adding `sin_wave` to the product

```{r}
wave <- (sin_wave * cos_wave) + sin_wave
```


Use the `plot()` function to plot your `wave`!

```{r}
plot(wave, type = "l")
```





# Indexing and Filtering Vectors


Create a vector `lotto_numbers` that contains the numbers 8, 26, 7, 20, 9, 0, 10

```{r}
lotto_numbers <- c(8, 26, 7, 20, 9, 0, 10)
```


Create a vector `prefix` that has the first three elements of `lotto_numbers`

```{r}
prefix <- lotto_numbers[1:3]
prefix
```

Create a vector `small` that has the values of `lotto_numbers` that are less than or equal to 10


```{r}
small <- lotto_numbers[lotto_numbers <= 10]
small
```


Create a vector `large` that has the values of `lotto_numbers` that are strictly greater than 10

```{r}
large <- lotto_numbers[lotto_numbers > 10]
large
```


Replace the values in `lotto_numbers` that are larger than 5 with the number 100

```{r}
lotto_numbers[lotto_numbers > 5] <- 100
lotto_numbers

```

Replace every odd-numbered value in `lotto_numbers` with the number 0

```{r}
lotto_numbers[lotto_numbers %% 2 == 1] <- 0
print(lotto_numbers)
```



